---
title: "novel-cell-types"
output: html_document
---

```{r, message=FALSE, warning=FALSE}
library("knitr")
library("BiocStyle")
library(monocle)
library(Hmisc)
BiocManager::install(c("monocle"))

devtools::install_github("kassambara/ggpubr")
library(ggpubr)


.cran_packages <- c("tidyverse", "Hmisc", "tsne")
.bioc_packages <- c("monocle")
.inst <- .cran_packages %in% installed.packages()

if(any(!.inst)) {
   install.packages(.cran_packages[!.inst])
}
.inst <- .bioc_packages %in% installed.packages()

if(any(!.inst)) {
   source("http://bioconductor.org/biocLite.R")
   biocLite(.bioc_packages[!.inst], ask = FALSE)
}
# Load packages into session, and print package version
sapply(c(.cran_packages, .bioc_packages), require, character.only = TRUE)

set.seed(12345)
```

```{r functions}
match_and_combine <- function(data1, col1, data2, col2) {
  d1 <- data1[(data1[,col1] %in% data2[,col2]), ]
  d2 <- data2[(data2[,col2] %in% data1[,col1]), ]
  
  m1 <- match(d1[,col1], d2[,col2])
  
  return(cbind(d1,d2[m1,]))
}

```

```{r}
load("../singleCell/Brian-Barron/GersteinResearch/18-07-19-Monocle.RData")
```

```{r}
x <- HSMM.reduced.clust@reducedDimA 
adf <- HSMM.reduced.clust@phenoData@data %>%
  data.frame

write.csv(plotdf, file = "~/Downloads/novel-cell-types_plotdf.csv")
```

```{r}
plotdf <- 
  x %>%
  t %>%
  data.frame %>% 
  cbind(., adf) %>%
  mutate(group = diseaseState) %>%
  mutate(group = gsub("[^asthma|control].*", "reference", group)) %>%
  mutate(group = capitalize(group)) %>%
  mutate(cluster.named = Cluster) %>%
  mutate(cluster.named = gsub(1, "B cells", Cluster)) %>%
  mutate(cluster.named = gsub(2, "NK cells", cluster.named)) %>%
  mutate(cluster.named = gsub(3, "Myeloid cells", cluster.named)) %>%
  mutate(cluster.named = gsub(4, "T cells", cluster.named))

p <-
  plotdf %>%
  ggplot(aes(x = X1, y = X2)) +
  geom_point(aes(color = Cluster)) +
  facet_wrap(~group) +
  labs(x = "t-sne 1",
       y = "t-sne 2") +
  theme_bw()

ann_text <- data.frame(X1 = c(8, -15, -12, 15), 
                       X2 = c(-10, -11, 5, 7),
                       group = factor("Reference", 
                                      levels = c("Asthma", "Control", "Reference")),
                       lab = c("B cells", "NK cells", "Myeloid cells", "T cells"))

p + geom_text(data = ann_text, aes(label = lab)) +
  ggsave("../figures/single-cell_facet-by-group.pdf", height = 3, width = 8)
```

Facet wrap by cluster, color by Asthma vs Control

```{r}
plotdf %>%
  ggplot(aes(X1, X2)) +
  geom_point(aes(color = group)) +
  facet_wrap(~cluster.named, scales = "free") +
  labs(x = "t-sne 1",
       y = "t-sne 2") +
  theme_bw() +
  ggsave("../figures/single-cell-facet-by-cluster.pdf", height = 4, width = 6)

```

# Figure 2C
## Boxplot of relative abundances of each cluster by patient

```{r}
df <- plotdf %>%
  filter(group %in% c("Asthma", "Control")) %>%
  group_by(batch, cluster.named, group) %>% 
  tally() %>%
  group_by(batch) %>%
  mutate(batch_total = sum(n)) %>%
  mutate(n.fraction = n / batch_total)

df <- 
  df %>%
  select(-n.fraction, -batch_total) %>%
  spread(cluster.named, n) %>%
  mutate(`B cells` = ifelse(is.na(`B cells`), 0, `B cells`)) %>%
  gather(-batch, -group, key = "cluster.named", value = "n") %>%
  group_by(batch) %>%
  mutate(batch_total = sum(n)) %>%
  mutate(n.fraction = n / batch_total)

wtpvals.greater <- list()
for (c in unique(df$cluster.named)) {
  wtpvals.greater[[c]] <- wilcox.test(n.fraction ~ group, data = df[df$cluster.named == c,], alternative = "greater")$p.value
}

mye.annot <- wtpvals.greater %>% unlist
```

```{r}
df %>%
  ggplot(aes(cluster.named, n.fraction)) +
  geom_boxplot(aes(fill = group)) +
  geom_signif(y_position = c(0.8), 
                xmin=c(1.8), 
                xmax = c(2.2),
                annotation = c(paste0("p-value = ", round(mye.annot[2], 2))), 
                tip_length = 0.02,
              textsize = 3.8) +
  theme_minimal() +
  labs(x = "Cluster",
       y = "Fraction of Cells") +
  ggsave("../figures/monocle-clusters_boxplot.pdf", height = 3, width = 6)
  
```


Which are signficantly different between asthmatics and controls? This will have a strong effect on the value of the interpretation. That is, if a new cell type is discovered that's significantly different between asthmatics and controls, that's more valuable in this context than a novel cell type that's the same in each.

An appropriate test for comparing two proportions is a z-test if it's parametric and fisher's exact test if non-parametric. The boxplots lead me to think non-parametric is most appropriate.

```{r}
# Create a 2x2 contigency table for each of the cell types in the single cell analysis
z <- 
  df %>%
  group_by(group, cluster.named) %>%
  dplyr::summarize(cells = sum(n),
                   total = sum(batch_total)) %>%
  mutate(not.cells = total - cells)

# 2x2 table for each cluster.named
ct <- list()    
for (c in z$cluster.named) {
   tmp <- z %>%
    filter(cluster.named == c) %>%
    select(group, cells, not.cells) %>%
    data.frame
    
   rownames(tmp) <- tmp$group
   ct[[c]] <- fisher.test(as.matrix(tmp[,-1]))$p.value
}

ct %>%
  unlist %>%
  p.adjust(., method = "fdr")
```

The myeloid cells are both significantly different between asthmatics and controls and visually have a lot of overlap. I'll see if I can split them into two clusters. 

```{r}
myeloid <- 
  plotdf %>%
  filter(cluster.named == "Myeloid cells")

myeloid <- 
  myeloid %>%
  select(X1, X2) %>%
  dist %>%
  hclust %>%
  cutree(., k = 2) %>%
  cbind(myeloid, .) %>%
  rename("new.cluster" = ".")
```

```{r}
myeloid %>%
  ggplot(aes(X1, X2)) +
  geom_point(aes(color = group, shape = factor(new.cluster)),
             size = 3) +
  theme_bw()
```

```{r}
myeloid.tally <- 
  myeloid %>%
  filter(group %in% c("Asthma", "Control")) %>%
  group_by(batch, new.cluster, group) %>% 
  tally() %>%
  group_by(batch) %>%
  mutate(batch_total = sum(n)) %>%
  mutate(n.fraction = n / batch_total)
  
myeloid.tally %>%
  ggplot() +
  geom_boxplot(aes(x = factor(new.cluster), y = n.fraction, fill = group)) +
  theme_bw() +
  labs(x = "Myeloid Subcluster",
       y = "Fraction of Cells") 
  # ggsave("../figures/monocle-clusters_boxplot.pdf", height = 3, width = 6)
  # ggplot(aes(X1, X2)) +
  # geom_point(aes(color = group, shape = factor(new.cluster)))
```

```{r}
# Create a 2x2 contigency table for each of the cell types in the single cell analysis
z2 <- 
  myeloid.tally %>%
  group_by(group, new.cluster) %>%
  dplyr::summarize(cells = sum(n),
                   total = sum(batch_total)) %>%
  mutate(not.cells = total - cells)

# 2x2 table for each cluster.named
ct2 <- list()    
for (c in z2$new.cluster) {
   tmp <- z2 %>%
    filter(new.cluster == c) %>%
    select(group, cells, not.cells) %>%
    data.frame
    
   rownames(tmp) <- tmp$group
   ct2[[c]] <- fisher.test(as.matrix(tmp[,-1]))$p.value
}

ct2 %>%
  unlist
```

Both are significantly different, though it seems to me that cluster 2 is more pronouncedly enriched in asthmatics.

# T cells split

```{r}
tcells <- 
  plotdf %>%
  filter(cluster.named == "T cells")

tcells <- 
  tcells %>%
  select(X1, X2) %>%
  dist %>%
  hclust %>%
  cutree(., k = 2) %>%
  cbind(tcells, .) %>%
  rename("new.cluster" = ".")
```

```{r}
tcells %>%
  ggplot(aes(X1, X2)) +
  geom_point(aes(color = group, shape = factor(new.cluster)),
             size = 3) +
  theme_bw() +
  
```

```{r}
tcells.tally <- 
  tcells %>%
  filter(group %in% c("Asthma", "Control")) %>%
  group_by(batch, new.cluster, group) %>% 
  tally() %>%
  group_by(batch) %>%
  mutate(batch_total = sum(n)) %>%
  mutate(n.fraction = n / batch_total)
  
tcells.tally %>%
  ggplot() +
  geom_boxplot(aes(x = factor(new.cluster), y = n.fraction, fill = group)) +
  theme_bw() +
  labs(x = "T cells Subcluster",
       y = "Fraction of Cells") 
  # ggsave("../figures/monocle-clusters_boxplot.pdf", height = 3, width = 6)
  # ggplot(aes(X1, X2)) +
  # geom_point(aes(color = group, shape = factor(new.cluster)))
```

```{r}
# Create a 2x2 contigency table for each of the cell types in the single cell analysis
z3 <- 
  tcells.tally %>%
  group_by(group, new.cluster) %>%
  dplyr::summarize(cells = sum(n),
                   total = sum(batch_total)) %>%
  mutate(not.cells = total - cells)

# 2x2 table for each cluster.named
ct3 <- list()    
for (c in z3$new.cluster) {
   tmp <- z3 %>%
    filter(new.cluster == c) %>%
    select(group, cells, not.cells) %>%
    data.frame
    
   rownames(tmp) <- tmp$group
   ct3[[c]] <- fisher.test(as.matrix(tmp[,-1]))$p.value
}

ct3 %>%
  unlist
```

No significant difference observed. I'll stick with the myeloid cell split. 

## Define the new cluster assignments

```{r}
myeloid.cluster <- 
  myeloid %>%
  mutate(myeloid.cluster = paste(Cluster, new.cluster, sep = ".")) %>%
  select(myeloid.cluster)

plotdf$mod.cluster <- as.character(plotdf$Cluster)
plotdf$mod.cluster[plotdf$Cluster == 3] <- myeloid.cluster$myeloid.cluster

```

